---
title: "Biostat 200C Homework 4"
author: "Cyrene Arputhasamy"
subtitle: Due May 12  @ 11:59PM
output: 
  html_document:
    toc: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE)
```

```{r}
library(faraway)
library(dplyr)
```

## Q1. ELMR Excercise 7.5 (p150)
The `debt` data arise from a large postal survey on the psychology of debt. The frequency of credit card use is a three-level factor ranging from `never`, through `occasionally` to `regularly`.

### 5a
Declare the response as an ordered factor and make a plot showing the relationship to `prodebt`. Comment on the plot. Use a table or plot to display the relationship between the response and the income group.
```{r}
debt$ccarduse <- ordered(debt$ccarduse, levels = c(1, 2, 3))
library(ggplot2)
ggplot(debt, aes(x = ccarduse, y = prodebt)) + geom_boxplot()
ggplot(debt, aes(x = ccarduse, y = incomegp)) + geom_boxplot()
debt %>%
  xtabs( ~ ccarduse + incomegp, .)
```
There is a little variability in `prodebt` across groups of card use. There is a general trend that those scoring high in `prodebt`, having a general favorable attitude to debt, score higher in credit card use. It also seems that those with high credit card use are generally in higher income groups. 

### 5b
Fit a proportional odds model for credit card use with all the other variables as predictors. What are the two most significant predictors (largest t-values) and what is their qualitative effect on the response? What is the least significant predictor?
```{r}
library(MASS)
pocc <- polr(ccarduse ~ ., data = debt)
summary(pocc)
exp(pocc$coef[1])
exp(pocc$coef[6])
```
The two most significant predictors are `incomegp` (income group) and `bankacc` (whether or not the respondent has a bank account). The least significant predictor is `house` (the security of housing tenure). The odds of moving from `never` to `occassional` credit card use **or** from `occasional` to `regularly` using a credit card increases by a factor of 1.6 when income group increases one level. The odds of moving from `never` to `occassional` credit card use **or** from `occasional` to `regularly` using a credit card increases by a factor of 8.18 when an individual has a bank account as compared to not having a bank account. 

### 5c
Fit a proportional odds model using only the least significant predictor from the previous model. What is the significance of this predictor in this small model? Are the conclusions regarding this predictor contradictory for the two models?
```{r}
pocc2 <- polr(ccarduse ~ house, data = debt)
summary(pocc2)
pt(3.895, 3, lower.tail = FALSE)
```
The significance of this predictor now has a p-value of 0.01 which is significant. This is contradictory to the previous model. 

### 5d
Use stepwise AIC to select a smaller model than the full set of predictors. You will need to handle the missing values carefully. Report on the qualitative effect of the predictors in your chosen model. Can we conclude that the predictors that were dropped from the model have no relation to the response?
```{r}
poccf <- polr(ccarduse ~ ., data = debt)
summary(poccf)
step(poccf, trace = FALSE)
```


### 5e
Compute the median values of the predictors in your selected model. At these median values, contrast the predicted outcome probabilities for both smokers and nonsmokers.

### 5f
Fit a proportional hazards model to the same set of predictors and recompute the two sets of probabilities from the previous question. Does it make a difference to use this type of model?

## Q2. Moments of exponential family distributions

Show that the exponential family distributions have moments
\begin{eqnarray*}
  \mathbb{E}Y &=& \mu = b'(\theta) \\
  \operatorname{Var}Y &=& \sigma^2 = b''(\theta) a(\phi).
\end{eqnarray*}

## Q3. Score and information matrix of GLM

Derive the gradient (score), negative Hessian, and Fisher information matrix (expected negative Hessian) of GLM.

GLM notes hw4 hint. the key is trying to understand the relationship between mu and sigma i squ. the systematic component eta equals xB. mu is a function of eta. in logistic regression mu = e^xb/1+e^xB. mu func of eta, eta a func of beta. take that into consideration when you do the derivative. in the hint i didn't take the cannonic link into consideration. these derivations are based on not using the cannonic link. 
der log likelihood w respect to beta. theta is one parameter in exp family. the first step is derivative w respect to beta but a(phi) is a constant and the second term is gone. def first w respect to theta, then der of theta w respect to beta. first w respect to theta, then beta. then need a relationship to have a direct relationship w beta. theta w respect to mu. my is a functino of eta which is a function of beta. 

taking the seconf deri gives.. the second term can be the same as the third term. use the relationship to take another deriv. sigma i sq so when you take another der by chair rule you can show it gives you another dsigma sq. take another def from the last term and plug into mu'' and get the same as the second term. 

## Q4. ELMR Exercise 8.1 (p171)

Data is generated from the exponential distribution with density $f(y) = \lambda e^{-\lambda y}$ where $\lambda, y > 0$. 

### 1a 
Identify the specific form of q,f,a(),b() and c() for the exponential distribu- tion.

### 1b
What is the canonical link and variance function for a GLM with a response following the exponential distribution?

### 1c
Identify a practical difficulty that may arise when using the canonical link in this instance.

### 1d
When comparing nested models in this case, should an F or c2 test be used? Explain.

### 1e
Express the deviance in this case in terms of the responses yi and the fitted values μˆi.


## Q5. ELMR Exercise 8.4 (p172)
Consider the Galápagos data and model analyzed in this chapter. The purpose of
this question is to reproduce the details of the GLM fitting of this data.

### 4a
Fit a Poisson model to the species response with the five geographic variables as predictors. Do not use the endemics variable. Report the values of the coef- ficients and the deviance.

### 4b
For a Poisson GLM, derive h, dh/dμ, V(μ) and the weights to be used in an iteratively fit GLM. What is the form of the adjusted dependent variable here?

### 4c
Using the observed response as initial values, compute the first stage of the iteration, stopping after the first linear model fit. Compare the coefficients of
this linear model to those found in the GLM fit. How close are they?

### 4d
Continue the iteration to get the next h and μ. Use this to compute the current
value of the deviance. How close is this to the deviance from the GLM?

### 4e
Compute one more iteration of the GLM fit, reporting the next calculation of
the coefficients and deviance. How close are these to target now?

### 4f
Repeattheseiterationsafewmoretimes,computingthedevianceineachtime. Stop when the deviance does not change much. Compare your final estimated
coefficients to that produced by the GLM fit.

### 4g
Use your final iterated linear model fit to produce standard errors for the coef-
ficients. How close are these to that produced by the direct GLM fit?

## Q6. ELMR Exercise 8.5 (p172)
Again using the Galápagos data, fit a Poisson model to the species response with the five geographic variables as predictors. Do not use the endemics variable. The purpose of this question is to compare six different ways of testing the significance of the elevation predictor, i.e., H0 : bElev = 0. In each case, report the p-value.

### 5a
Use the z-statistic from the model summary.

### 5b
Fit a model without elevation and use the difference in deviances to make the
test.

### 5c
Use the Pearson Chi-squared statistic in place of the deviance in the previous
test.

### 5d
Fit the Poisson model with a free dispersion parameter as described in Sec-
tion 5.2. Make the test using the model summary.

### 5e
Use the sandwich estimation method for the standard errors in the original
model. Use these to compute z-statistics.

### 5f
Use the robust GLM estimation method and report the test result from the
summary.

### 5g
Compare all six results. Pick the best one and justify your choice.