---
title: "Biostat 200C Homework 2"
subtitle: Due Apr 28 @ 11:59PM
output: 
  html_document:
    toc: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE)
```

To submit homework, please upload both Rmd and html files to Bruinlearn by the deadline.
```{r}
sessionInfo()
```
```{r}
library(faraway)
library(dplyr)
```

## Q1. CFR of COVID-19

Of primary interest to public is the risk of dying from COVID-19. A commonly used measure is case fatality rate/ratio/risk (CFR), which is defined as
$$
\frac{\text{number of deaths from disease}}{\text{number of diagnosed cases of disease}}.
$$
Apparently CFR is not a fixed constant; it changes with time, location, and other factors. Also CFR is different from the infection fatality rate (IFR), the probability that someone infected with COVID-19 dies from it. 

In this exercise, we use logistic regression to study how US county-level CFR changes according to demographic information and some health-, education-, and economy-indicators.

### Data sources

- `04-04-2020.csv.gz`: The data on COVID-19 confirmed cases and deaths on 2020-04-04 is retrieved from the [Johns Hopkins COVID-19 data repository](https://github.com/CSSEGISandData/COVID-19). It was downloaded from this [link](https://github.com/CSSEGISandData/COVID-19) (commit 0174f38). This repository has been archived by the owner on Mar 10, 2023. It is now read-only. You can download data from box: <https://ucla.box.com/s/brb3vz4nwoq8pjkcutxncymqw583d39l>

- `us-county-health-rankings-2020.csv.gz`: The 2020 County Health Ranking Data was released by [County Health Rankings](https://www.countyhealthrankings.org). The data was downloaded from the [Kaggle Uncover COVID-19 Challenge](https://www.kaggle.com/roche-data-science-coalition/uncover) (version 1). You can download data from box: <https://ucla.box.com/s/brb3vz4nwoq8pjkcutxncymqw583d39l>

### Sample code for data preparation

Load the `tidyverse` package for data manipulation and visualization.
```{r}
# tidyverse of data manipulation and visualization
library(tidyverse)
```
Read in the data of COVID-19 cases reported on 2020-04-04.
```{r}
county_count <- read_csv("./datasets/04-04-2020.csv.gz") %>%
  # cast fips into dbl for use as a key for joining tables
  mutate(FIPS = as.numeric(FIPS)) %>%
  filter(Country_Region == "US") %>%
  print(width = Inf)
```
Standardize the variable names by changing them to lower case.
```{r}
names(county_count) <- str_to_lower(names(county_count))
```
Sanity check by displaying the unique US states and territories:
```{r}
county_count %>%
  dplyr::select(province_state) %>%
  distinct() %>%
  arrange(province_state) %>%
  print(n = Inf)
```
We want to exclude entries from `Diamond Princess`, `Grand Princess`, `Guam`, `Northern Mariana Islands`, `Puerto Rico`, `Recovered`, and `Virgin Islands`, and only consider counties from 50 states and DC.
```{r}
county_count <- county_count %>%
  filter(!(province_state %in% c("Diamond Princess", "Grand Princess", 
                                 "Recovered", "Guam", "Northern Mariana Islands", 
                                 "Puerto Rico", "Virgin Islands"))) %>%
  print(width = Inf)
```
Graphical summarize the COVID-19 confirmed cases and deaths on 2020-04-04 by state.
```{r}
county_count %>%
  # turn into long format for easy plotting
  pivot_longer(confirmed:recovered, 
               names_to = "case", 
               values_to = "count") %>%
  group_by(province_state) %>%
  ggplot() + 
  geom_col(mapping = aes(x = province_state, y = `count`, fill = `case`)) + 
  # scale_y_log10() + 
  labs(title = "US COVID-19 Situation on 2020-04-04", x = "State") + 
  theme(axis.text.x = element_text(angle = 90))
```

Read in the 2020 county-level health ranking data.
```{r}
county_info <- read_csv("./datasets/us-county-health-rankings-2020.csv.gz") %>%
  filter(!is.na(county)) %>%
  # cast fips into dbl for use as a key for joining tables
  mutate(fips = as.numeric(fips)) %>%
  dplyr::select(fips, 
         state,
         county,
         percent_fair_or_poor_health, 
         percent_smokers, 
         percent_adults_with_obesity, 
         # food_environment_index,
         percent_with_access_to_exercise_opportunities, 
         percent_excessive_drinking,
         # teen_birth_rate, 
         percent_uninsured,
         # primary_care_physicians_rate,
         # preventable_hospitalization_rate,
         # high_school_graduation_rate,
         percent_some_college,
         percent_unemployed,
         percent_children_in_poverty,
         # `80th_percentile_income`,
         # `20th_percentile_income`,
         percent_single_parent_households,
         # violent_crime_rate,
         percent_severe_housing_problems,
         overcrowding,
         # life_expectancy,
         # age_adjusted_death_rate,
         percent_adults_with_diabetes,
         # hiv_prevalence_rate,
         percent_food_insecure,
         # percent_limited_access_to_healthy_foods,
         percent_insufficient_sleep,
         percent_uninsured_2,
         median_household_income,
         average_traffic_volume_per_meter_of_major_roadways,
         percent_homeowners,
         # percent_severe_housing_cost_burden,
         population_2,
         percent_less_than_18_years_of_age,
         percent_65_and_over,
         percent_black,
         percent_asian,
         percent_hispanic,
         percent_female,
         percent_rural) %>%
  print(width = Inf)
```

For stability in estimating CFR, we restrict to counties with $\ge 5$ confirmed cases.
```{r}
county_count <- county_count %>%
  filter(confirmed >= 5)
```
We join the COVID-19 count data and county-level information using FIPS (Federal Information Processing System) as key. 
```{r}
county_data <- county_count %>%
  left_join(county_info, by = "fips") %>%
  print(width = Inf)
```
Numerical summaries of each variable:
```{r}
summary(county_data)
```
List rows in `county_data` that don't have a match in `county_count`:
```{r}
county_data %>%
  filter(is.na(state) & is.na(county)) %>%
  print(n = Inf)
```
We found there are some rows that miss `fips`. 
```{r}
county_count %>%
  filter(is.na(fips)) %>%
  dplyr::select(fips, admin2, province_state) %>%
  print(n = Inf)
```
We need to (1) manually set the `fips` for some counties, (2) discard those `Unassigned`, `unassigned` or `Out of`, and (3) try to join with `county_info` again.
```{r}
county_data <- county_count %>%
  # manually set FIPS for some counties
  mutate(fips = ifelse(admin2 == "DeKalb" & province_state == "Tennessee", 47041, fips)) %>%
  mutate(fips = ifelse(admin2 == "DeSoto" & province_state == "Florida", 12027, fips)) %>%
  #mutate(fips = ifelse(admin2 == "Dona Ana" & province_state == "New Mexico", 35013, fips)) %>% 
  mutate(fips = ifelse(admin2 == "Dukes and Nantucket" & province_state == "Massachusetts", 25019, fips)) %>% 
  mutate(fips = ifelse(admin2 == "Fillmore" & province_state == "Minnesota", 27045, fips)) %>%  
  #mutate(fips = ifelse(admin2 == "Harris" & province_state == "Texas", 48201, fips)) %>%  
  #mutate(fips = ifelse(admin2 == "Kenai Peninsula" & province_state == "Alaska", 2122, fips)) %>%  
  mutate(fips = ifelse(admin2 == "LaSalle" & province_state == "Illinois", 17099, fips)) %>%
  #mutate(fips = ifelse(admin2 == "LaSalle" & province_state == "Louisiana", 22059, fips)) %>%
  #mutate(fips = ifelse(admin2 == "Lac qui Parle" & province_state == "Minnesota", 27073, fips)) %>%  
  mutate(fips = ifelse(admin2 == "Manassas" & province_state == "Virginia", 51683, fips)) %>%
  #mutate(fips = ifelse(admin2 == "Matanuska-Susitna" & province_state == "Alaska", 2170, fips)) %>%
  mutate(fips = ifelse(admin2 == "McDuffie" & province_state == "Georgia", 13189, fips)) %>%
  #mutate(fips = ifelse(admin2 == "McIntosh" & province_state == "Georgia", 13191, fips)) %>%
  #mutate(fips = ifelse(admin2 == "McKean" & province_state == "Pennsylvania", 42083, fips)) %>%
  mutate(fips = ifelse(admin2 == "Weber" & province_state == "Utah", 49057, fips)) %>%
  filter(!(is.na(fips) | str_detect(admin2, "Out of") | str_detect(admin2, "Unassigned"))) %>%
  left_join(county_info, by = "fips") %>%
  print(width = Inf)
```
Summarize again
```{r}
summary(county_data)
```
If there are variables with missing value for many counties, we go back and remove those variables from consideration.

Let's create a final data frame for analysis.
```{r}
county_data <- county_data %>%
  mutate(state = as.factor(state)) %>%
  dplyr::select(county, confirmed, deaths, state, percent_fair_or_poor_health:percent_rural)
summary(county_data)
```
Display the 10 counties with highest CFR.
```{r}
county_data %>%
  mutate(cfr = deaths / confirmed) %>%
  dplyr::select(county, state, confirmed, deaths, cfr) %>%
  arrange(desc(cfr)) %>%
  top_n(10)
```
Write final data into a csv file for future use.
```{r}
write_csv(county_data, "./datasets/covid19-county-data-20200404.csv.gz")
```

### Note:

Given that the datasets were collected in the middle of the pandemic, what assumptions of CFR might be violated by defining CFR as `deaths/confirmed` from this data set? 

Because COVID-19 pandemic was still ongoing in 2020, we should realize some critical assumptions for defining CFR are not met using this dataset.

1. Numbers of confirmed cases do not reflect the number of diagnosed people. This is mainly limited by the availability of testing.

2. Some confirmed cases may die later.

With acknowledgement of these severe limitations, we continue to use `deaths/confirmed` as a very rough proxy of CFR.

### Q1.1 

Read and run above code to generate a data frame `county_data` that includes county-level COVID-19 confirmed cases and deaths, demographic, and health related information. 

### Q1.2 

What assumptions of logistic regression may be violated by this data set?

One of the assumptions of logistic regression is that the observations are independent from one another. In this case since the observations are counties and cases of COVID, they may not be independent in counties that are adjacent to one another.

### Q1.3 

Run a logistic regression, using variables `state`, ..., `percent_rural` as predictors. 

```{r}
lmod <- glm(cbind(deaths, confirmed - deaths) ~ . - county, family = binomial, data = county_data)
summary(lmod)
```


### Q1.4

Interpret the regression coefficients of 3 significant predictors with p-value <0.01.

- A one percent increase in population less than 18 years of age decreases the odds of death among confirmed cases of COVID by roughly 5% on average (exp(-4.812e-02) = 0.9530), holding all else constant in the state of Alabama. 

- A one percent increase in fair or poor health in Alabama increases the odds of death among confirmed cases of COVID by .3% on average (exp(0.003092) = 1.003), holding all else constant.

- COVID cases occurring in the District of Columbia have a .9% increase in odds of death as compared to other states holding all else constant. 

### Q1.5 

Apply analysis of deviance to (1) evaluate the goodness of fit of the model and (2) compare the model to the intercept-only model. 

```{r}
pchisq(lmod$deviance, lmod$df.residual, lower = FALSE)
```

The model is not a good fit as the p-value is less than .05. This does **not** mean that the predictors are insignificant of the odds of death out of COVID cases in a county -- rather, it means our model has been "miss-calibrated" or "miss-specified" in some way. For example, it could mean that we are missing many other important predictors in our model, or it could mean that a model assumption has not been reasonably satisfied.

### Q1.6

Perform analysis of deviance to evaluate the significance of each predictor. Display the 10 most significant predictors.

```{r}
aod <- drop1(lmod, test = "Chi")
arrange(aod, `Pr(>Chi)`) %>%
  head(10)
```
The top ten most significant predictors are `state`, `percent_homeowners`, `percent_insufficient_sleep`, `percent_severe_housing_problems`, `percent_children_in_poverty`, `median_household_income`, `percent_some_college`, `percent_fair_or_poor_health`, `percent_less_than_18_years_of_age`, and `percent_smokers`. 

### Q1.7

Construct confidence intervals of regression coefficients.
```{r}
library(gtsummary)
lmod %>%
  tbl_regression(intercept = TRUE, exponentiate = TRUE) %>%
  suppressWarnings()
# library(MASS)
# confint(lmod)
```


### Q1.8 

Plot the deviance residuals against the fitted values. Are there potential outliers?
```{r}
county_data %>%
    mutate(devres = residuals(lmod, type = "deviance"),
    linpred = predict(lmod, type = "link")) %>%
    ggplot +
        geom_point(mapping = aes(x = linpred, y = devres)) +
        labs(x = "Linear predictor", y = "Deviance residual")
```
It looks like there is on potential outlier with a residual less than -5.0 and a group of observations that all fall around deviance residual of 0.

### Q1.9

Plot the half-normal plot. Are there potential outliers in predictor space?

```{r}
halfnorm(hatvalues(lmod))
```
```{r}
county_data %>%
  slice(367, 931)
```
There are two observations in the predictor space that are potential outliers. The counties of District of Columbia and New York both have a percent with access to exercise opportunities of 100% which is unusual. It also might be unusual that these counties are 0% rural which might make them outliers in the predictor space.

### Q1.10

Find the best sub-model using the AIC criterion.
```{r}
step(lmod, trace = TRUE)
```
AIC Criterion dropped 7 variables. 

### Q1.11

Find the best sub-model using the lasso with cross validation.
glmnet doesn't have the functionality for binomial but has glm for bernoulli. you have to transform the dataset into the format w obs_wt the way we did in binomial notes o-ring. then you can use glmnet and lasso. outcome vector as outcome.

```{r}
obs_wt = c(rbind(county_data$deaths, county_data$confirmed - county_data$deaths))
county_data_long <- county_data %>%
  slice(rep(1:n(), each = 2)) %>% # replicate each row twice
  mutate(deaths = rep(c(1, 0), 1446)) %>%
  mutate(obs_wt = obs_wt)
county_data_long %>% print(width = Inf)

glm(deaths ~ . - county - confirmed - obs_wt, weights = obs_wt, family = binomial, data = county_data_long) %>%
summary()
```


## Q2. Odds ratios

Consider a $2 \times 2$ contingency table from a prospective study in which people who were or were not exposed to some pollutant are followed up and, after several years, categorized according to the presence or absence of a disease. Following table shows the probabilities for each cell. The odds of disease for either exposure group is $O_i = \pi_i / (1 - \pi_i)$, for $i = 1,2$, and so the odds ratio is
$$
\phi = \frac{O_1}{O_2} = \frac{\pi_1(1 - \pi_2)}{\pi_2 (1 - \pi_1)}
$$
is a measure of the relative likelihood of disease for the exposed and not exposed groups.

|             | Diseased | Not diseased |
|:-----------:|----------|--------------|
| Exposed     | $\pi_1$  | $1 - \pi_1$  |
| Not exposed | $\pi_2$  | $1 - \pi_2$  |

### Q2.1

For the simple logistic model
$$
\pi_i = \frac{e^{\beta_i}}{1 + e^{\beta_i}}, 
$$
show that if there is no difference between the exposed and not exposed groups (i.e., $\beta_1 = \beta_2$), then $\phi = 1$.

**Solution:** 

If there is no difference between the exposed and not exposed groups, then $\beta_1 = \beta_2$. Then, 

\begin{eqnarray*}
\pi_1 $=$ \frac{e^{\beta_1}}{1 + e^{\beta_1}} $=$ \frac{e^{\beta_2}}{1 + e^{\beta_2}} $=$ \pi_2 \\
\pi_1 $=$ \pi_2
\end{eqnarray*}

Then, 

\begin{eqnarray*}
\phi $=$ \frac{O_1}{O_2} $=$ \frac{\pi_1(1 - \pi_2)}{\pi_2 (1 - \pi_1)} \\
$=$ \frac{\pi_1(1 - \pi_1)}{\pi_1 (1 - \pi_1)} $=$ 1
\end{eqnarray*}

### Q2.2 

Consider $J$ $2 \times 2$ tables, one for each level $x_j$ of a factor, such as age group, with $j=1,\ldots, J$. For the logistic model
$$
\pi_{ij} = \frac{e^{\alpha_i + \beta_i x_j}}{1 + e^{\alpha_i + \beta_i x_j}}, \quad i = 1,2, \quad j= 1,\ldots, J.
$$
Show that $\log \phi$ is constant over all tables if $\beta_1 = \beta_2$.

## Q3. ELMR Chapter 4 Excercise 3

Data on social mobility of men in the UK may be found in `cmob`. A sample of men aged 45–64 was drawn from the 1971 census and 1981 census and the social class of the man was recorded at each timepoint. The classes are I = professional, II = semiprofessional, IIIN = skilled non-manual, HIM = skilled manual, IV = semiskilled, V = unskilled.

### Q3.1 Check for symmetry, quasi-symmetry, marginal homogeneity and quasi- independence.

We start by checking **independence**:
```{r}
cmob %>%
  print(width = Inf)
tab <- xtabs(cmob) %>% print()
summary(tab)
```
It seems that there's strong evidence of dependence, which isn't surprising since mobility of social class probably depends strongly on the starting point. Now we can check for **symmetry**. We fit such a model by defining a factor where the levels represent the symmetric pairs of the off-diagonal elements.
```{r}
(symfac <- factor(apply(cmob[,2:3], 1,
                        function(x) paste (sort(x), collapse="-"))))
mods <- glm(y ~ symfac, cmob, family=poisson)
c(deviance(mods), df.residual(mods))
pchisq(deviance(mods), df.residual(mods), lower = F)
```
Here we see evidence of a lack of symmetry, so we can check the residuals. 
```{r}
round(xtabs(residuals(mods) ~ class71 + class81, cmob),3)
```
We see that there are more negative residuals above the diagonal, while those below the diagonal are mostly positive. So there is generally more unskilled in '71 who moved up by '81 than the reverse. We can go on to check **homogeneity**:
```{r}
margin.table(tab, 1)
margin.table(tab, 2)
```
We can see that there are somewhat more professional-class males in '81 than in '71, and more unskilled males in '71 than in '81 so perhaps marginal homogeneity does not hold here. The assumption of symmetry implies marginal homogeneity but the reverse is not necessarily true. The **quasi-symmetry** model allows for some symmetry while allowing for different marginals. We can fit the quasi-symmetry model:
```{r}
modq <- glm(y ~ class71 + class81 + symfac, cmob, family = poisson)
pchisq(deviance(modq), df.residual(modq), lower = F)
```
This model does not fit either. We can test for **marginal homogeneity** by comparing the symmetry and quasi-symmetry models:
```{r}
anova(mods, modq, test = "Chi")
```
We find evidence of marginal homogeneity but it doesn't hold since we did not find quasi-symmetry. 

We also notice that there are many people who fall in the diagonal of the original cross tabulation showing the number of people who did not change social status. We might ask whether there is independence between social class in '71 and social class in '81 among those people whose class did not change. This is the **quasi-independence** hypothesis which we can test by omitting the data from the diagonal:
```{r}
modqi <- glm(y ~ class71 + class81, cmob, family=poisson, 
             subset=-c(1, 8, 15, 22, 29, 36))
pchisq(deviance(modqi), df.residual(modqi), lower = F)
```
The model does not fit. This is not surprising since we can see that the groups close to the diagonal are larger than those farther away. The difference in social class between the two years is likely to be smaller than expected under independence. 

### Q3.2 Develop a score-based model. Find some good-fitting scores.

First we can fit a nominal-by-nominal model:
```{r}
nomod <- glm(y ~ class71 + class81, cmob, family = poisson)
pchisq(deviance(nomod), df.residual(nomod), lower = F)
```
We find evidence against independence. However, we can take advantage of the ordinal structure of the class variables and define some scores. Since there's no reason for the contrary, we assign evenly spaced scores:
```{r}
cmob$class71o <- unclass(cmob$class71)
cmob$class81o <- unclass(cmob$class81)
ormod <- glm(y ~ class71 + class81 + I(class71o*class81o), cmob, family= poisson)
anova(nomod, ormod, test="Chi")
```
We see that there is some evidence of an association. Using the ordinal information gives us more power to detect the association. 
We can examine $\hat{\gamma}$:
```{r}
summary(ormod)$coef['I(class71o*class81o)',]
```
To assess the robustness of the assignment of scores we can assign other scores so that there is more distinction between higher class and lower class. 
```{r}
aclass <- c(1,4,5,6,9)
bclass <- c(1,1,5,9,9)
ormoda <- glm(y ~ class71 + class81 + I(aclass[class71o]*bclass[class81o]), cmob, family= poisson)
anova(nomod, ormoda, test="Chi")
```

