---
title: "Biostat 200C Homework 5"
subtitle: Due June 2  @ 11:59PM
author: "Cyrene Arputhasamy"
output: 
  html_document:
    toc: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE)
```
## Q1. Balanced one-way ANOVA random effects model

Consider the balanced one-way ANOVA random effects model with $a$ levels and $n$ observations in each level
$$
y_{ij} = \mu + \alpha_i + \epsilon_{ij}, \quad i=1,\ldots,a, \quad j=1,\ldots,n.
$$
where $\alpha_i$ are iid from $N(0,\sigma_\alpha^2)$, $\epsilon_{ij}$ are iid from $N(0, \sigma_\epsilon^2)$. 

1. Derive the ANOVA estimate for $\mu$, $\sigma_\alpha^2$, and $\sigma_{\epsilon}^2$. Specifically show that
\begin{eqnarray*}
  \mathbb{E}(\bar y_{\cdot \cdot}) &=& \mathbb{E} \left( \frac{\sum_{ij} y_{ij}}{na} \right) = \mu \\
  \mathbb{E} (\text{SSE}) &=& \mathbb{E} \left[ \sum_{i=1}^a \sum_{j=1}^n (y_{ij} - \bar{y}_{i \cdot})^2 \right] = a(n-1) \sigma_{\epsilon}^2 \\
  \mathbb{E} (\text{SSA}) &=& \mathbb{E} \left[ \sum_{i=1}^a \sum_{j=1}^n (\bar{y}_{i \cdot} - \bar{y}_{\cdot \cdot})^2 \right] = (a-1)(n \sigma_{\alpha}^2 + \sigma_{\epsilon}^2),
\end{eqnarray*}
which can be solved to obtain ANOVA estimate
\begin{eqnarray*}
\widehat{\mu} &=& \frac{\sum_{ij} y_{ij}}{na}, \\
\widehat{\sigma}_{\epsilon}^2 &=& \frac{\text{SSE}}{a(n-1)}, \\
\widehat{\sigma}_{\alpha}^2 &=& \frac{\text{SSA}/(a-1) - \widehat{\sigma}_{\epsilon}^2}{n}.
\end{eqnarray*}

**Solution:**
+ $\mathbb{E}(\bar y_{\cdot \cdot})$
\begin{eqnarray*}
\mathbb{E}(y_{ij}) &=& \mathbb{E}(\mu + \alpha_i + \epsilon_{ij}) \\
&=& \mathbb{E}\mu + \mathbb{E}\alpha_i + \mathbb{E}\epsilon_{ij} \\
&=& \mu + 0 + 0 = \mu \\
\\
\mathbb{E}\frac{\sum_{ij}y_{ij}}{na} &=& \frac{\sum_{ij} \mathbb{E}(y_{ij})}{na} \\
&=& \frac{\sum_{ij}\mu}{na} = \mu
\end{eqnarray*}

+ $\widehat{\sigma}_{\epsilon}^2 &=& \frac{\text{SSE}}{a(n-1)}$
\begin{eqnarray*}
\mathbb{E} \left[ \sum_{i=1}^a \sum_{j=1}^n (y_{ij} - \bar{y}_{i \cdot})^2 \right] &=& \sum_{i=1}^a \mathbb{E}\left[ \sum_{j=1}^n (y_{ij} - \bar{y}_{i \cdot})^2 \right] \\
y_{ij} = \mu + \alpha_i+\epsilon_{ij} \\
\bar{y}_{i \cdot} &=& \frac{1}{n} \sum_{j=1}^n(\mu + \alpha_i + \epsilon_{ij}) = \mu + \alpha_i + \bar{\epsilon}_{i \cdot} \\
y_{ij} - \bar{y}_{i \cdot} &=& \epsilon_{ij} - \bar{\epsilon}_{i \cdot} \\
\mathbb{E}(\sum(\epsilon_{ij} - \bar{\epsilon}_{i \cdot})^2) = (n-1)\sigma_\epsilon^2 \\
\sum_{i=1}^a\mathbb{E}\left[\sum_{j=1}^n(\epsilon_{ij} - \bar{\epsilon}_{i \cdot})^2 \right] = \sum^a (n-1)\sigma_\epsilon^2 \\
&=& a(n-1)\sigma_\epsilon^2
\end{eqnarray*}
2. Derive the MLE estimate for $\mu$, $\sigma_\alpha^2$, and $\sigma_{\epsilon}^2$. Hint: write down the log-likelihood and find the maximizer.

3. (**Optional**) Derive the REML estimate for $\mu$, $\sigma_\alpha^2$, and $\sigma_{\epsilon}^2$. 

4. For all three estimates, check that your results match those we obtained using R for the `pulp` example in class.

## Q2. Estimation of random effects

1. Assume the conditional distribution
$$
\mathbf{y} \mid \boldsymbol{\gamma} \sim N(\mathbf{X} \boldsymbol{\beta} + \mathbf{Z} \boldsymbol{\gamma}, \sigma^2 \mathbf{I}_n)
$$
and the prior distribution
$$
\boldsymbol{\gamma} \sim N(\mathbf{0}_q, \boldsymbol{\Sigma}).
$$
Then by the Bayes theorem, the posterior distribution is
\begin{eqnarray*}
f(\boldsymbol{\gamma} \mid \mathbf{y}) &=& \frac{f(\mathbf{y} \mid \boldsymbol{\gamma}) \times f(\boldsymbol{\gamma})}{f(\mathbf{y})}, \end{eqnarray*}
where $f$ denotes corresponding density. Show that the posterior distribution is a multivariate normal with mean
$$
\mathbb{E} (\boldsymbol{\gamma} \mid \mathbf{y}) = \boldsymbol{\Sigma} \mathbf{Z}^T (\mathbf{Z} \boldsymbol{\Sigma} \mathbf{Z}^T + \sigma^2 \mathbf{I})^{-1} (\mathbf{y} - \mathbf{X} \boldsymbol{\beta}).
$$

2. For the balanced one-way ANOVA random effects model, show that the posterior mean of random effects is always a constant (less than 1) multiplying the corresponding fixed effects estimate.


## Q3. ELMR Exercise 11.1 (p251)
```{r}
library(faraway)
library(dplyr)
```

The `ratdrink` data consist of five weekly measurements of body weight for 27 rats. The first 10 rats are on a control treatment while 7 rats have thyroxine added to their drinking water. Ten rats have thiouracil added to their water.
```{r}
ratdrink %>% print(width = Inf)
```


### 11.1a
Plot the data showing how weight increases with age on a single panel, taking care to distinguish the three treatment groups. Now create a three-panel plot, one for each group. Discuss what can be seen.
```{r}
ggplot(data = ratdrink, 
       aes(x = weeks, y = wt, group = treat, col = as.factor(treat))) + 
  stat_summary(geom = "line", fun.y = mean) + 
  theme_pubr() + 
  labs(x = "Week", y = "Weight",
                      title = "Rat Weight Over Time",
                      col = "Treatment") +
  scale_color_manual(values = c("control" = "red", "thyroxine" = "green",
                                "thiouracil" = "blue"),
                     labels = c("Control", "Thyroxine", 
                                "Thiouracil")) +
  theme(legend.position = "right")
controlp <- ratdrink %>%
  filter(treat == "control") %>%
  ggplot(aes(x = weeks, y = wt)) + geom_point() +
  labs(title = "Control Rat Weight Over Time")
thyp <- ratdrink %>%
  filter(treat == "thyroxine") %>%
  ggplot(aes(x = weeks, y = wt)) + geom_point() + 
  labs(title = "Thyroxine Rat Weight Over Time")
thip <- ratdrink %>%
  filter(treat == "thiouracil") %>%
  ggplot(aes(x = weeks, y = wt)) + geom_point() +
  labs(title = "Thiouracil Rat Weight Over Time")
grid.arrange(controlp, thyp, thip, ncol = 2)
```
Control mice and thiouracil treated mice have a similar weight trajectory over 4 weeks while thyroxine treated mice seem to have a different and lower trajectory, ending with an average weight visibly different from the other mice. 

### 11.1b
Fit a linear longitudinal model that allows for a random slope and intercept for each rat. Each group should have a different mean line. Give interpretation for the following estimates:

 + The fixed effect intercept term.
 + The interaction between thiouracil and week
 + The intercept random effect SD.
 
```{r}
library(tidyverse)
```
```{r}
m1 <- lmer(wt ~ weeks * treat + (weeks | subject), data = ratdrink)
summary(m1)
```
+ The fixed effect intercept term : Similarly to GLM, the fixed effects intercept indicates that the average weight of control mice is 52.8 (no units in data sheet) on average.

+ The interaction between `thiouracil` and `weeks` : The weight of mice goes down by 9 in thiouracil treated mice as compared to control mice dependent on weeks. 

+ The intercept random effect SD : The variability between subjects within a treatment group is 5.7, the random effect intercept SD. 
 
### 11.1c 
Check whether there is a significant treatment effect.
```{r}
library(pbkrtest)
m2 <- lmer(wt ~ weeks + (weeks | subject), data = ratdrink)
KRmodcomp(m1, m2)
```
The p-value of the F-test indicates that there is a significant effect of the treatment.

### 11.1d
Construct diagnostic plots showing the residuals against the fitted values and a QQ plot of the residuals. Interpret.
```{r}
plot(resid(m1) ~ fitted(m1), xlab = "Fitted", ylab = "Residuals")
abline(h = 0 )
```
The residuals are mostly scattered about the x-axis with no evidence of homoscedasticity, therefore the assumptions of linearity and equal variance appear reasonably satisfied.

```{r}
qqnorm(resid(m1), main = "")
```
The QQ plot shows that there don't seem to be deviations from the normality assumption.

### 11.1e
Construct confidence intervals for the parameters of the model. Which random effect terms may not be significant? Is the thyroxine group significantly different from the control group?
```{r}
confint(m1, method = "boot")
```
$\sigma_1$ is the variance of the random intercept, $\sigma_2$ is the covariance of the random intercept and random slope, and $\sigma_3$ is the variance of the random slope. We aren't concerned with $\sigma_2$. The intervals for both $\sigma_1$ and $\sigma_3$ don't include $0$ so they are significant and can remain in the model. The `thyroxine` group might not be significantly different from the `control` group since the interval includes 0.

## Q4. ELMR Exercise 13.1 (p295)
The `ohio` data concern 536 children from Steubenville, Ohio and were taken as part of a study on the effects of air pollution. Children were in the study for 4 years from ages 7 to 10. The response was whether they wheezed or not. 

The variables are:
+ `resp` an indicator of wheeze status (1 = yes, 0 = no)
+ `id` an identifier for the child
+ `age` 7yrs = -2, 8yrs = -1, 9yrs = 0, 10yrs = 1
+ `smoke` an indicator of maternal smoking at the first year of the study (1 = smoker, 0 = nonsmoker)
```{r}
ohio %>% print(width = Inf)
```

### 13.1a
Do any of the mothers in the study change their smoking status during the period of observation?
```{r}
table(ohio$smoke)
ohio %>%
  dplyr::select(id, smoke) %>%
  n_distinct()
```
There are 537 children, so there should be 537 unique combinations of `id` and `smoke` if the mothers did not change their smoking status. Since there are 537 unique combinations, that means no mothers changed her smoking status.

### 13.1b
Construct a table that shows proportion of children who wheeze for 0, 1, 2, 3 or 4 years broken down by maternal smoking status.
```{r}
xtabs(~ smoke + resp, ohio)
library(gtsummary)
ohio %>%
  tbl_strata(strata = smoke,
             .tbl_fun = ~ .x %>%
  tbl_cross(row = resp, col = age, percent = "cell")) %>%
  bold_labels()
```

### 13.1c
Make plot which shows how the proportion of children wheezing changes by age with a separate line for smoking and nonsmoking mothers.
```{r}
props <- xtabs(~ age + resp + smoke, ohio) %>%
  as_tibble() 
props$np <- props$n/537
props
props %>%
  filter(resp == 1) %>%
  ggplot() +
  geom_line(aes(x = age, y = np, group = smoke, col = as.factor(smoke))) +
  theme_pubr() + labs(x = "Age", y = "Proportion",
                      title = "Proportion of Child Wheezing by Age",
                      col = "Mother Smoker")
  scale_color_manual(values = c("0" = "red", "1" = "green"),
                     labels = c("Mother not a smoker", "Mother smokes"))
```

### 13.1d
(d) Group the data by child to count the total (out of four) years of wheezing. Fit a binomial GLM to this response to check for a maternal smoking effect. Does this prove there is a smoking effect or could there be another plausible explanation?
```{r}
(df1 <- aggregate(ohio$smoke, list(ohio$id), FUN=sum) %>%
  as_data_frame() %>%
  dplyr::rename("mom_smoke" = "x", "id" = "Group.1"))

(ohiobin <- aggregate(ohio$resp, list(ohio$id), FUN=sum) %>%
  mutate(not_wheeze = 4 - x) %>%
  dplyr::rename("years_wheezed" = "x", "id" = "Group.1") %>%
  full_join(df1, by = "id"))
ohiobin$smoke <- ifelse(ohiobin$mom_smoke != 0, 1, 0)
ohiobin

glm(cbind(years_wheezed, not_wheeze) ~ smoke, family = binomial, data = ohiobin) %>%
  summary()
```



### 13.1e
(e) Fit a model for each individual response using a GLMM fit using penalized quasi-likelihood. Describe the effects of age and maternal smoking. How do the odds of wheezing change numerically over time?
```{r}
library(MASS)
pqlmod <- glmmPQL(resp ~ age + smoke, random = ~ 1 | id, family = binomial, data = ohio)
summary(pqlmod)
```


### 13.1f
(f) Now fit the same model but using adaptive Gaussian-Hermit quadrature. Compare to the previous model fit.
```{r}
library(lme4)
ghmod <- glmer(resp ~ age + smoke + (1 | id), nAGQ = 25, family = binomial, data = ohio)
summary(ghmod)
```


### 13.1g OPTIONAL
(g) Use INLA to fit the same model. What does this model say about the effect of age and maternal smoking?

### 13.1h OPTIONAL
(h) Use STAN to fit the same model. Check the MCMC diagnostics and again dis- cuss the age and maternal smoking effects.

### 13.1i
(i) Fit the model using GEE. Use an autoregressive rather than exchangeable error structure. Compare the results to the previous model fits. In your model, what indicates that a child who already wheezes is likely to continue to wheeze?
```{r}
library(geepack)
geemod <- geeglm(resp ~ age + smoke, id = id, corstr = "ar1", scale.fix = TRUE, data = ohio, family =binomial(link = "logit"))
summary(geemod)
```

### 13.1j
(j) What is your overall conclusion regarding the effect of age and maternal smoking? Can we trust the GLM result or are the GLMM models preferable?